Sample-efficient Reinforcement Learning \\ with Implicitly Quantized Representations

### 2024-02-01 paper/main.tex
added algorithm

% if you use cleveref..
\usepackage[capitalize,noabbrev,nameinlink]{cleveref}

@@

% \newcommand{\our}{\textsc{vq-td3}\xspace}
\newcommand{\our}{\textsc{iFSQ-RL}\xspace}

@@

In this section, we detail our method, titled Vector Quantized TD3 (\our).
\our is conceptually simple, it first learns a representation of the observation space and then,
given this representation, it performs model-free RL (e.g. TD3) on this representation.
% It the first representation learning phase, it learns an encoder which maps observations $o \in \mathcal{O}$
% to a latent space $z \in \mathcal{Z}$.
% It first learns a representation of the observation space and then uses
See \cref{alg:main_alg}.

We consider Markov Decision Processes (MDPs, \citet{bellmanMarkovianDecisionProcess1957a}) $\mathcal{M} = (\mathcal{O}, \mathcal{A}, \mathcal{P}, \mathcal{R}, \gamma,\mathcal{T})$, where an agent receives
observation $o_{t} \in \mathcal{O}$ at time step $t$ and performs an action $a_{t} \in \mathcal{A}$.
The agent then obtains a reward $r_{t} = \mathcal{R} (o_{t}, a_{t})$ and obtains the next observation
$o_{t+1} = P(\cdot \mid o_{t}, a_{t})$.
discount factor $\gamma \in [0, 1]$

\our has four parameterized components.
\begin{algorithm}[tb]
   \caption{\our}
   \label{alg:main_alg}
\begin{algorithmic}
   \STATE {\bfseries Input:} Encoder $e_{\theta}$, transition $d_{\phi}$, critics $Q_{\psi}$, policy $\pi_{\eta}$
   \FOR{$i$ {\bfseries to} $N_{\text{episodes}}$}
    \STATE $\mathcal{D} \leftarrow \mathcal{D} \cup \{o_{t}, a_{t}, o_{t+1}, r_{t+1}\}^{T}_{t=0}$
    \FOR{$i=1$ {\bfseries to} $T \times UTD$}
        \STATE Representation update
        \STATE TD3 update
    \ENDFOR
   \ENDFOR
\end{algorithmic}
\end{algorithm}

\begin{align}
&\text{Encoder: } & z_{t} &= e_{\theta} (s_{t}) \label{eq:encoder} \\
&\text{Dynamics: } & \hat{z}_{t+1} &= d_{\phi} (z_{t}, a_{t}) \label{eq:transition} \\
&\text{Value: } & q_{t} &= Q_{\psi} (z_{t}, a_{t}) \label{eq:value} \\
&\text{Policy: } & a_{t} &\sim \pi_{\eta} (z_{t}) , \label{eq:policy}
\end{align}
Note that unlike TCRL \citep{zhaoSimplifiedTemporalConsistency2023}, our transition model does
not predict the reward.

\textbf{Representation learning}
Our representation learning is based soley on the temporal consistency loss,
However, unlike previous works, we normalize the latent vector
Our representation learning consists of an encoder $e_{\theta}$ (which maps observations to
latent states) and a transition model $d_{\phi}$ which maps
Our representation learning adopts a self-supervised loss.

\begin{align} \label{eq:rep-loss}
  \mathcal{L}_{\text{rep}}(\theta, \phi; \mathcal{D})
 % &= \E_{s_{t} \sim \mathcal{D}} \left[ \sum_{h=0}^{H-1} \| \hat{z}_{t+1} - z_{t+1} \|_{2}^{2} \right] \\
%&= \E_{(s_{t}, a_{t}, s_{t+1}) \sim \mathcal{D}}
&= \E_{\tau \sim \mathcal{D}}
\left[ \sum_{h=0}^{H-1} \gamma^{h} \| d_{\phi}(e_{\theta}(s_{t}), a_{t}) - e_{\bar{\theta}}(s_{t+1}) \|_{2}^{2} \right]
\end{align}

\textbf{Finite Scalar Quantization}
We follow Finite Scalar Quantization (FSQ, \cite{mentzerFiniteScalarQuantization2023})
and bound each dimension of our latent space.
Given the output of our MLP encoder $z \in \R^{d}$,

>>>>

% if you use cleveref..
% \usepackage[capitalize,noabbrev,nameinlink]{cleveref}
\usepackage[capitalize,nameinlink]{cleveref}

@@

% \newcommand{\our}{\textsc{vq-td3}\xspace}
% \newcommand{\our}{\textsc{iFSQ-RL}\xspace}
\newcommand{\our}{\textsc{iQRL}\xspace}

@@

In this section, we detail our method, titled \textit{implicitly Quantized Reinforcement Learning} (\our).
\our is conceptually simple, it (1) learns a representation of the observation space and then,
(2) performs model-free RL (e.g. TD3) on this representation.
% It the first representation learning phase, it learns an encoder which maps observations $o \in \mathcal{O}$
% to a latent space $z \in \mathcal{Z}$.
% It first learns a representation of the observation space and then uses
See \cref{fig:diagram-self-predictive} and \cref{alg:main_alg}.

We consider Markov Decision Processes (MDPs, \citet{bellmanMarkovianDecisionProcess1957a}) $\mathcal{M} = (\mathcal{O}, \mathcal{A}, \mathcal{P}, \mathcal{R}, \gamma)$,
where an agent receives an observation $o_{t} \in \mathcal{O}$ at time step $t$ and performs an action $a_{t} \in \mathcal{A}$.
The agent then obtains a reward $r_{t} = \mathcal{R} (o_{t}, a_{t})$ and obtains the next observation
$o_{t+1} = P(\cdot \mid o_{t}, a_{t})$.
The discount factor is denoted $\gamma \in [0, 1]$.

\begin{algorithm}[tb]
   \caption{\our}
   \label{alg:main_alg}
   \renewcommand{\algorithmiccomment}[1]{\hfill\textcolor{gray}{\(\triangleright\) #1}}
\begin{algorithmic}
   \STATE {\bfseries Input:} Encoder $e_{\theta}$, transition $d_{\phi}$, critics $\{Q_{\psi_{1}}, Q_{\psi_{2}} \}$, policy $\pi_{\eta}$, learning rate $\alpha$, target network update rate $\tau$
   \FOR{$i$ {\bfseries to} $N_{\text{episodes}}$}
    \STATE Collect data in environment
    \STATE $\qquad \mathcal{D} \leftarrow \mathcal{D} \cup \{o_{t}, a_{t}, o_{t+1}, r_{t+1}\}^{T}_{t=0}$
    \FOR{$i=1$ {\bfseries to} $T$}
        \STATE Update representation
        \STATE $\qquad [\theta, \phi] \leftarrow [\theta, \phi] + \alpha \nabla \left( \mathcal{L}_{\text{rep}}(\theta, \phi; \mathcal{D}) \right)$  \COMMENT{\cref{eq:rep-loss}}
        \STATE Update critic
        \STATE \quad \quad $\psi \leftarrow \psi + \alpha \nabla \left( \mathcal{L}_{Q}(\psi; \mathcal{D}) \right)$ \COMMENT{\cref{eq:value-loss}}
        \IF{i \% 2 == 0}
          \STATE Update actor less frequently than critic
          \STATE $\quad \quad \eta \leftarrow \psi + \alpha \nabla \left( \mathcal{L}_{\pi}(\psi; \mathcal{D}) \right)$  \COMMENT{\cref{eq:policy-loss}}
          % \STATE $\bar{\eta} \leftarrow \tau \bar{\eta} + (1-\tau) \eta$
        \ENDIF
        \STATE Update target networks
        \STATE $\quad \quad [\bar{\theta}, \bar{\phi}, \bar{\psi}] \leftarrow \tau [\bar{\theta}, \bar{\phi}, \bar{\psi}] + (1-\tau) [{\theta}, {\phi}, {\psi}]$
    \ENDFOR
   \ENDFOR
\end{algorithmic}
\end{algorithm}

\our has four parameterized components.
\begin{align}
&\text{Encoder: } & z_{t} &= e_{\theta} (s_{t}) \label{eq:encoder} \\
&\text{Dynamics: } & \hat{z}_{t+1} &= d_{\phi} (z_{t}, a_{t}) \label{eq:transition} \\
&\text{Value: } & q_{t} &= Q_{\psi} (z_{t}, a_{t}) \label{eq:value} \\
&\text{Policy: } & a_{t} &\sim \pi_{\eta} (z_{t}) , \label{eq:policy}
\end{align}
Note that unlike TCRL \citep{zhaoSimplifiedTemporalConsistency2023}, our transition model does
not predict the reward.

\textbf{Representation learning}
Our representation learning is based soley on the temporal consistency loss,
However, unlike previous works, we normalize the latent vector
Our representation learning consists of an encoder $e_{\theta}$ (which maps observations to
latent states) and a transition model $d_{\phi}$ which maps
Our representation learning adopts a self-supervised loss.
%
\begin{align} \label{eq:rep-loss}
  \mathcal{L}_{\text{rep}}(\theta, \phi; \tau)
 % &= \E_{s_{t} \sim \mathcal{D}} \left[ \sum_{h=0}^{H-1} \| \hat{z}_{t+1} - z_{t+1} \|_{2}^{2} \right] \\
%&= \E_{(s_{t}, a_{t}, s_{t+1}) \sim \mathcal{D}}
% &= \E_{\tau \sim \mathcal{D}}
% \left[ \sum_{h=0}^{H-1} \gamma^{h} \| d_{\phi}(e_{\theta}(s_{t}), a_{t}) - e_{\bar{\theta}}(s_{t+1}) \|_{2}^{2} \right]
&= \sum_{h=0}^{H-1} \gamma^{h} \| d_{\phi}(e_{\theta}(s_{t}), a_{t}) - e_{\bar{\theta}}(s_{t+1}) \|_{2}^{2}
\end{align}
%
\textbf{Finite Scalar Quantization}
We follow Finite Scalar Quantization (FSQ, \cite{mentzerFiniteScalarQuantization2023})
and bound each dimension of our latent space.
Given the output of our MLP encoder $z \in \R^{d}$,
